Supplemental Material: Multidimensional analysis and detection of informative features in human brain white matter

### 2021-04-23 reports/manuscript/paper.tex

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}
\usepackage[capitalize]{cleveref}
\usepackage[inline]{enumitem}
\usepackage[per-mode=symbol,group-separator={,},group-minimum-digits=4]{siunitx}
\usepackage{listings}
\usepackage{todonotes}
\usepackage{subcaption}
\usepackage{url}

@@

\replaced{\mbox{\cite{stejskal1965spin, wandell2016clarifying}}}{\mbox{\cite{wandell2016clarifying}}}.

@@

samples along the length of each bundle \replaced{\mbox{{\cite{Jones2005-yq, yeatman2012tract, colby2012, ODonnell2009-uu}}}}{\mbox{\cite{yeatman2012tract,
colby2012, ODonnell2009-uu}}},

@@

\replaced{\mbox{\cite{colby2012, Nichols2002-zu, Nichols2003-yy, chamberland2019dimensionality}}}{\mbox{\cite{colby2012, Nichols2002-zu, Nichols2003-yy}}}.

@@

One more approach to multivariate analysis of neuroimaging data is to perform inference or prediction in a transformed feature space rather then the original diffusion metric feature space. For example, in Lasso PCR, diffusion features can be projected onto a principal components (PC) basis for use in a sparsity constrained model \mbox{\cite{powell2018local, rasero2021integrating}. While these approaches achieve high predictive performance and may even yield biologically interpretable principal components \mbox{\cite{chamberland2019dimensionality}}, they neglect anatomical grouping information. We view these approaches as complementary to the SGL-based approach: PCR based approaches seeks to model brain-phenotype relationships using the most parsimonious representation of variance in the diffusion measures. In contrast, our SGL-based approach seeks to establish whether prior knowledge of anatomical grouping improves modeling of brain-phenotype relationships. Motivated by this distinction, we also introduce a union of the two approaches called PCR-SGL, in which each group of features is independently transformed into its PC basis, thereby retaining the anatomical grouping information of the original feature space. We demonstrate circumstances in which this approach both helps and hinders predictive performance.}

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL and PCR-SGL also identify
the white matter tracts most important for ALS classification.
\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \mbox{\cite{sarica2017corticospinal}.}
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups (see supplemental Figures S1, S2 for tract profiles of all 18 tracts in these groups).
We find that SGL and PCR-SGL select FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

@@

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the tract profile
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets see supplemental Figures S3-S8 for the full tract profile information in all tracts/datasets.

@@

In contrast to the ALS classification case, the selected $\alpha$ values
indicate high global sparsity over group sparsity, with $\alpha = $
{\whLRatio}, {\hbnLRatio}, and {\ccLRatio}, for the WH, HBN, and Cam-CAN
datasets, respectively. The model weights are distributed over many different
tracts and dMRI tissue properties (see
\cref{fig:age-results:wh-coefs3d,fig:age-results:hbn-coefs3d,fig:age-results:cc-coefs3d}
and supporing Figures in \nameref{S1_Appendix}).
This demonstrates that SGL is not coerced to produce overly sparse results
when a more accurate model requires a dense selection of features.
Furthermore, inspecting the portions of bundles with larger coefficients in
\cref{fig:age-results:wh-profile,fig:age-results:hbn-profile,fig:age-results:cc-profile}
reveals that SGL selects informative regions where diffusion properties are
different between the age quintiles.

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.
In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item different acquisition parameters, which can be challenging to harmonize \mbox{\cite{pinto2020harmonization}}
    \item different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets
    \item different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets
    \item different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).
\end{enumerate*}
In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \mbox{\cite{lermausabiaga2019replication}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.}

@@

Reconstruction was performed using \emph{QSIprep} 0.12.1. Multi-tissue fiber response functions were estimated using the dhollander algorithm. FODs were estimated via constrained spherical deconvolution (CSD, \mbox{\cite{originalcsd, tournier2008csd}) using an unsupervised multi-tissue method \mbox{\cite{dhollander2019response, dhollander2016unsupervised}}. Reconstruction was done using MRtrix3 \mbox{\cite{mrtrix3}}. FODs were intensity-normalized using mtnormalize \mbox{\cite{mtnormalize}}.}
\end{itemize}

@@

These processes create bundle profiles, in which diffusion measures
are quantified and averaged along eighteen major fiber tracts,
which are enumerated in the supplemental material.
See Figure S3 of \mbox{\cite{kruper2021evaluating} for a depiction of these white matter bundles.}
Here,
we use only the mean diffusivity (MD) and the fractional anisotropy
(FA) of the diffusion tensor, but additional dMRI-based maps or maps
based on other quantitative MRI measurements can also be used.
The resulting feature space was the same for all four datasets, with the FA and MD metrics at each of 100 nodes in eighteen bundles comprising 3600 features per subject.
These bundle profiles, along with the phenotypical data we wish to explain
or predict, form the input to the SGL algorithm. In a domain-agnostic
machine learning context, the phenotypical data comprise the target
variables while the bundle profiles form the feature or predictor
variables (see \cref{fig:methods:group-structure}).

@@

For the multisite HBN study, we use the ComBat harmonization method to robustly adjust for site effects in the tract profiles. Initially designed to correct for site effects in gene expression studies \mbox{\cite{johnson2007adjusting}, ComBat employs a parametric empirical Bayes approach to adjust for batch effects and has since been applied to multi-site cortical thickness measurements \mbox{\cite{fortin2018harmonization}}, multi-site DTI studies \mbox{\cite{fortin2017harmonization}}, and functional MRI data in the Adolescent Brain Cognitive Development Study (ABCD) \mbox{\cite{nielson2018detecting}}. We rely on the \texttt{neurocombat\_sklearn} library \mbox{\cite{neurocombat_sklearn}}, to apply ComBat in our scikit-learn analysis pipeline and present bundle profile site differences in the supplemental material.}

@@

Often, the target variable $y$ is not in the domain in which the linear
model can be best fit to it. \Cref{eq:lm-approx} can be slightly
modified as:
\begin{equation}
    \hat{y} = f^{-1} \left( \mathbf{X} \hat{\beta} \right),
    \label{eq:lm-transform}
\end{equation}
where the transformation function $f^{-1}$ characterizes the transform
applied to the data before fitting the linear coefficients.
This is similar to the use of a link function in a generalized linear model, but without the adoption of an exponential probability distribution \mbox{\cite{nelder1972technique}.}
For the WH, HBN, and Cam-CAN datasets, we use a
logarithmic transform,
\begin{equation}
    f \left( \hat{y} \right) = \ln \left( \hat{y} \right),
    \label{eq:log-nonlinearity}
\end{equation}
implemented using scikit-learn's \lstinline|TransformedTargetRegressor| meta-estimator \mbox{\cite{scikit-learn} with numpy's \lstinline|log| and \lstinline|exp| as the transform and inverse transform functions, respectively \mbox{\cite{harris2020array}}.}

>>>>

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}
\usepackage[capitalize]{cleveref}
\crefformat{equation}{Eq~(#2#1#3)}
\crefrangeformat{equation}{Eqs~(#3#1#4) to~(#5#2#6)}
\crefmultiformat{equation}{Eqs~(#2#1#3)}%
{ and~(#2#1#3)}{, (#2#1#3)}{ and~(#2#1#3)}
\crefformat{figure}{Fig~#2#1#3}
\crefrangeformat{figure}{Figs~#3#1#4 to~#5#2#6}
\crefmultiformat{figure}{Figs~#2#1#3}%
{ and~#2#1#3}{, #2#1#3}{ and~#2#1#3}

\usepackage[inline]{enumitem}
\usepackage[per-mode=symbol,group-separator={,},group-minimum-digits=4]{siunitx}
\usepackage{listings}
\usepackage{todonotes}
\usepackage{subcaption}
\usepackage{url}

@@

\cite{stejskal1965spin, wandell2016clarifying}.

@@

samples along the length of each bundle
\cite{Jones2005-yq, yeatman2012tract, colby2012, ODonnell2009-uu},

@@

\cite{colby2012, Nichols2002-zu, Nichols2003-yy, chamberland2019dimensionality}.

@@

One more approach to multivariate analysis of neuroimaging data is to perform inference or prediction in a transformed feature space rather then the original diffusion metric feature space. For example, in Lasso PCR, diffusion features can be projected onto a principal components (PC) basis for use in a sparsity constrained model \cite{powell2018local, rasero2021integrating}. While these approaches achieve high predictive performance and may even yield biologically interpretable principal components \cite{chamberland2019dimensionality}, they neglect anatomical grouping information. We view these approaches as complementary to the SGL-based approach: PCR based approaches seeks to model brain-phenotype relationships using the most parsimonious representation of variance in the diffusion measures. In contrast, our SGL-based approach seeks to establish whether prior knowledge of anatomical grouping improves modeling of brain-phenotype relationships. Motivated by this distinction, we also introduce a union of the two approaches called PCR-SGL, in which each group of features is independently transformed into its PC basis, thereby retaining the anatomical grouping information of the original feature space. We demonstrate circumstances in which this approach both helps and hinders predictive performance.

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL and PCR-SGL also identify
the white matter tracts most important for ALS classification.
\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \cite{sarica2017corticospinal}.
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups (see supplemental Figs S1 and S2 for tract profiles of all 18 tracts in these groups).
We find that SGL and PCR-SGL select FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

@@

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the tract profile
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets see supplemental Figs S3 to S8 for the full tract profile information in all tracts/datasets.

@@

In contrast to the ALS classification case, the selected $\alpha$ values
indicate high global sparsity over group sparsity, with $\alpha = $
{\whLRatio}, {\hbnLRatio}, and {\ccLRatio}, for the WH, HBN, and Cam-CAN
datasets, respectively. The model weights are distributed over many different
tracts and dMRI tissue properties (see
\cref{fig:age-results:wh-coefs3d,fig:age-results:hbn-coefs3d,fig:age-results:cc-coefs3d}
and supporing figures in \nameref{S1_Appendix}).
This demonstrates that SGL is not coerced to produce overly sparse results
when a more accurate model requires a dense selection of features.
Furthermore, inspecting the portions of bundles with larger coefficients in
\cref{fig:age-results:wh-profile,fig:age-results:hbn-profile,fig:age-results:cc-profile}
reveals that SGL selects informative regions where diffusion properties are
different between the age quintiles.

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.
In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item different acquisition parameters, which can be challenging to harmonize \cite{pinto2020harmonization}
    \item different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets
    \item different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets
    \item different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).
\end{enumerate*}
In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \cite{lermausabiaga2019replication}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.

@@

Reconstruction was performed using \emph{QSIprep} 0.12.1. Multi-tissue fiber response functions were estimated using the dhollander algorithm. FODs were estimated via constrained spherical deconvolution (CSD, \cite{originalcsd, tournier2008csd}) using an unsupervised multi-tissue method \cite{dhollander2019response, dhollander2016unsupervised}. Reconstruction was done using MRtrix3 \cite{mrtrix3}. FODs were intensity-normalized using mtnormalize \cite{mtnormalize}.
\end{itemize}

@@

These processes create bundle profiles, in which diffusion measures
are quantified and averaged along eighteen major fiber tracts,
which are enumerated in the supplemental material.
See Fig S3 of \cite{kruper2021evaluating} for a depiction of these white matter bundles.
Here,
we use only the mean diffusivity (MD) and the fractional anisotropy
(FA) of the diffusion tensor, but additional dMRI-based maps or maps
based on other quantitative MRI measurements can also be used.
The resulting feature space was the same for all four datasets, with the FA and MD metrics at each of 100 nodes in eighteen bundles comprising 3600 features per subject.
These bundle profiles, along with the phenotypical data we wish to explain
or predict, form the input to the SGL algorithm. In a domain-agnostic
machine learning context, the phenotypical data comprise the target
variables while the bundle profiles form the feature or predictor
variables (see \cref{fig:methods:group-structure}).

@@

For the multisite HBN study, we use the ComBat harmonization method to robustly adjust for site effects in the tract profiles. Initially designed to correct for site effects in gene expression studies \cite{johnson2007adjusting}, ComBat employs a parametric empirical Bayes approach to adjust for batch effects and has since been applied to multi-site cortical thickness measurements \cite{fortin2018harmonization}, multi-site DTI studies \cite{fortin2017harmonization}, and functional MRI data in the Adolescent Brain Cognitive Development Study (ABCD) \cite{nielson2018detecting}. We rely on the \texttt{neurocombat\_sklearn} library \cite{neurocombat_sklearn}, to apply ComBat in our scikit-learn analysis pipeline and present bundle profile site differences in the supplemental material.

@@

Often, the target variable $y$ is not in the domain in which the linear
model can be best fit to it. \Cref{eq:lm-approx} can be slightly
modified as:
\begin{equation}
    \hat{y} = f^{-1} \left( \mathbf{X} \hat{\beta} \right),
    \label{eq:lm-transform}
\end{equation}
where the transformation function $f^{-1}$ characterizes the transform
applied to the data before fitting the linear coefficients.
This is similar to the use of a link function in a generalized linear model, but without the adoption of an exponential probability distribution \cite{nelder1972technique}.
For the WH, HBN, and Cam-CAN datasets, we use a
logarithmic transform,
\begin{equation}
    f \left( \hat{y} \right) = \ln \left( \hat{y} \right),
    \label{eq:log-nonlinearity}
\end{equation}
implemented using scikit-learn's \lstinline|TransformedTargetRegressor| meta-estimator \cite{scikit-learn} with numpy's \lstinline|log| and \lstinline|exp| as the transform and inverse transform functions, respectively \cite{harris2020array}.

### 2021-04-23 reports/manuscript/paper_changes.tex

samples along the length of each bundle \replaced{\mbox{{\cite{Jones2005-yq, yeatman2012tract, colby2012, ODonnell2009-uu}}}}{\mbox{\cite{yeatman2012tract,
colby2012, ODonnell2009-uu}}},

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
\added{We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL \added{and PCR-SGL} also \replaced{identify}{identifies}
the white matter tracts most important for ALS classification.
\added{\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \mbox{\cite{sarica2017corticospinal}}.}
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups \added{(see supplemental Figures S1, S2 for tract profiles of all 18 tracts in these groups)}.
We find that SGL \replaced{and PCR-SGL select}{selects} FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

@@

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the \replaced{tract profile}{tractometry-extracted}
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets \added{see supplemental Figures S3-S8 for the full tract profile information in all tracts/datasets}.

@@

These processes create bundle profiles, in which diffusion measures
are quantified and averaged along eighteen major fiber tracts,
which are enumerated in the supplemental material.
\added{See Figure S3 of \mbox{\cite{kruper2021evaluating}} for a depiction of these white matter bundles.}
Here,
we use only the mean diffusivity (MD) and the fractional anisotropy
(FA) of the diffusion tensor, but additional dMRI-based maps or maps
based on other quantitative MRI measurements can also be used.
\added{The resulting feature space was the same for all four datasets, with the FA and MD metrics at each of 100 nodes in eighteen bundles comprising 3600 features per subject.}
These bundle profiles, along with the phenotypical data we wish to explain
or predict, form the input to the SGL algorithm. In a domain-agnostic
machine learning context, the phenotypical data comprise the target
variables while the bundle profiles form the feature or predictor
variables (\replaced{see}{See} \cref{fig:methods:group-structure}).

>>>>

samples along the length of each bundle
\replaced{\mbox{\cite{Jones2005-yq, yeatman2012tract, colby2012, ODonnell2009-uu}},}{\mbox{\cite{yeatman2012tract, colby2012, ODonnell2009-uu}},}

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
\added{We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL \added{and PCR-SGL} also \replaced{identify}{identifies}
the white matter tracts most important for ALS classification.
\added{\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \mbox{\cite{sarica2017corticospinal}}.}
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups \added{(see supplemental Figs S1 and S2 for tract profiles of all 18 tracts in these groups)}.
We find that SGL \replaced{and PCR-SGL select}{selects} FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

@@

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the \replaced{tract profile}{tractometry-extracted}
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets \added{see supplemental Figs S3 to S8 for the full tract profile information in all tracts/datasets}.

@@

These processes create bundle profiles, in which diffusion measures
are quantified and averaged along eighteen major fiber tracts,
which are enumerated in the supplemental material.
\added{See Fig S3 of \mbox{\cite{kruper2021evaluating}} for a depiction of these white matter bundles.}
Here,
we use only the mean diffusivity (MD) and the fractional anisotropy
(FA) of the diffusion tensor, but additional dMRI-based maps or maps
based on other quantitative MRI measurements can also be used.
\added{The resulting feature space was the same for all four datasets, with the FA and MD metrics at each of 100 nodes in eighteen bundles comprising 3600 features per subject.}
These bundle profiles, along with the phenotypical data we wish to explain
or predict, form the input to the SGL algorithm. In a domain-agnostic
machine learning context, the phenotypical data comprise the target
variables while the bundle profiles form the feature or predictor
variables (\replaced{see}{See} \cref{fig:methods:group-structure}).

### 2021-04-23 reports/manuscript/supplemental.tex

>>>>

\section{Data harmonization in the HBN dataset}

\Cref{fig:hbn-site-bp:fa,fig:hbn-site-bp:md} show the bundle profiles for the HBN dataset separated by scanner site: Rutgers Rutgers University Brain Imaging Center (RU) and the CitiGroup Cornell Brain Imaging Center (CBIC). We see strong site differences in both FA and MD profiles. Conversely, \cref{fig:hbn-site-age-dist} shows that the age distributions at each site are similar. ComBat effectively harmonizes the scanner site differences, as shown in \cref{fig:hbn-harmonized-bp:fa,fig:hbn-harmonized-bp:md}.

\begin{figure}
    \includegraphics[width=0.9\textwidth]{hbn_site_age_distributions.pdf}
    \caption{%
        {%
            \bf Age distributions are similar between the different
            HBN sites.
        }
        \label{fig:hbn-site-age-dist}
    }
\end{figure}

\begin{figure}
    \includegraphics[width=0.9\textwidth]{hbn_site_profiles_fa.pdf}
    \caption{%
        {%
            \bf Fractional anisotropy (FA) bundle profiles
            exhibit strong site differences in the HBN dataset.
        }
        \label{fig:hbn-site-bp:fa}
    }
\end{figure}

\begin{figure}
    \includegraphics[width=0.9\textwidth]{hbn_site_profiles_md.pdf}
    \caption{%
        {%
            \bf Mean diffusivity (MD) bundle profiles
            exhibit strong site differences in the HBN dataset.
        }
        \label{fig:hbn-site-bp:md}
    }
\end{figure}

\begin{figure}
    \includegraphics[width=0.9\textwidth]{hbn_harmonized_profiles_fa.pdf}
    \caption{%
        {%
            \bf Site differences in fractional anisotropy (FA)
            do not survive ComBat harmonization.
        }
        \label{fig:hbn-harmonized-bp:fa}
    }
\end{figure}

\begin{figure}
    \includegraphics[width=0.9\textwidth]{hbn_harmonized_profiles_md.pdf}
    \caption{%
        {%
            \bf Site differences in mean diffusivity (MD)
            do not survive ComBat harmonization.
        }
        \label{fig:hbn-harmonized-bp:md}
    }
\end{figure}

\pagebreak
